---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
This is my second attempt at the Reproducible Research MOOC.  
Data was cloned to my working directory using git bash.

### Loading and preprocessing the data

```{r, echo=TRUE}
activity <-read.csv("activity.csv")
head(activity)
summary(activity)
```

### What is mean total number of steps taken per day?

- Calculate the total number of steps taken per day
- (ignore the missing values in the dataset)
```{r, echo=TRUE}
totalsteps <- aggregate(steps~date, data=activity, FUN=sum, na.rm=TRUE)
```

- Make Histogram of the total number of steps taken each day
```{r, echo=TRUE}
hist(totalsteps$steps, 
    col = "red", 
    main = "Total Number of Steps Taken Each Day", 
    xlab="Number of Steps (Day)", 
    breaks = 15)
```

- Calculate and report the mean and median of the total number of steps taken 
per day 
```{r, echo=TRUE}
meansteps <- mean(totalsteps$steps, na.rm=TRUE, scientific=FALSE)
mediansteps <- median(totalsteps$steps, na.rm=TRUE)
```

The mean total number of steps per day is `r meansteps` 
The median total number of steps per day is `r mediansteps`

### What is the average daily activity pattern?
- Make a time series plot of the 5-minute interval (x-axis) and the average 
number of steps taken, averaged across all days (y-axis)

```{r, echo=TRUE}
averageDaily <- aggregate(steps~interval, data=activity, FUN=mean)
plot(averageDaily, 
    type = "l", 
    col = "red",
    xlab = "5-minute interval", 
    ylab = "average number of steps taken")

maxStep <- averageDaily[which.max(averageDaily$steps),]$interval
```

The 5-minute interval, on average across all the days in the dataset, that 
contains the maximum number of steps is `r maxStep`.

### Imputing missing values
there are a number of days/intervals where there are missing values (NA). 
The presence of missing days may introduce bias into some calculations or 
summaries of the data.

Devise a strategy for filling in all of the missing values in the dataset. 
The strategy does not need to be sophisticated. For example, you could use 
the mean/median for that day, or the mean for that 5-minute interval, etc.

- Calculate and report the total number of missing values in the dataset 
(i.e. the total number of rows with NAs)

```{r, echo=TRUE} 
missingValues <- sum(is.na(activity))
```

The number of missing values in the data set is `r missingValues`

- Create a new dataset that is equal to the original dataset but with the 
missing data filled in.  
```{r, echo=TRUE} 
imputedActivity<-activity
imputedActivity[is.na(imputedActivity[, 1]), 1] <-averageDaily[is.na(imputedActivity[, 1]),2]
imputedtotalsteps <- aggregate(steps~date, data=imputedActivity, FUN=sum)
```

Make a histogram of the total number of steps taken each day
```{r, echo=TRUE}
hist(imputedtotalsteps$steps, 
     col = "red",
     breaks = 15, 
     xlab = "Imputed total number of steps (day)", 
     main = "Total Imputed steps each day" )
```

- Calculate and report the mean and median total number of steps taken per day. 
```{r, echo=TRUE} 
imputedmeansteps <- mean(imputedtotalsteps$steps)
imputedmediansteps <- median(imputedtotalsteps$steps)

```

The imputed mean is `r imputedmeansteps`.
The imputed median is `r imputedmediansteps`.   
There was no change in the values of the mean and median from the first part of 
the assignment.

### Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. 
Use the dataset with the filled-in missing values for this part.

- Create a new factor variable in the dataset with two levels - "weekday" 
and "weekend" indicating whether a given date is a weekday or weekend day.

```{r, echo=TRUE}
library(lattice)
library(plyr)
weekdays <- weekdays(as.Date(imputedActivity$date))
data_weekdays <- transform(imputedActivity, day=weekdays)
data_weekdays$wk <- ifelse(data_weekdays$day %in% c("Saturday", "Sunday"),
                           "weekend", "weekday")
average_week <- ddply(data_weekdays, .(interval, wk), summarise, steps=mean(steps))
```

- Make a panel plot containing a time series plot (i.e. type = "l") of the 
5-minute interval (x-axis) and the average number of steps taken, averaged 
across all weekday days or weekend days (y-axis). 
```{r, echo=TRUE}
xyplot(steps ~ interval | wk, aggregate(steps ~ interval + wk, data_weekdays, 
                                        FUN=mean), layout = c(1, 2), type="l")                              
```

There are differences in activity patterns between the weekdays and the weekend.
The weekends have a flatter profile with higher numbers of more steps taken.